---
title: Module 2 - Chi-Square Testing and ANOVA
author: "Justin Ehringhaus"
date: "`r format(Sys.time(), '%B %d, %Y')`"
output: 
  github_document:
    toc: true
    toc_depth: 5
bibliography: "Ehringhaus-references.bib"
nocite: '@*'
---

```{r setup, include=FALSE}
# repo: https://github.com/ehringhaus/regression-diagnostics
knitr::opts_chunk$set(echo = TRUE)
```

---

```{r echo=FALSE}
# █▀█ █▀▀ █▀ █▀▀ ▀█▀ █▀
# █▀▄ ██▄ ▄█ ██▄ ░█░ ▄█
# clear console
cat("\014")
# clear global environment
rm(list = ls())
# clear plots
try(dev.off(dev.list()["RStudioGD"]), silent = TRUE)
# clear packages
try(p_unload(p_loaded(), character.only = TRUE), silent = TRUE)
# disables scientific notion for entire R session
options(scipen = 100)
```

```{r}
# █▀█ ▄▀█ █▀▀ █▄▀ ▄▀█ █▀▀ █▀▀ █▀
# █▀▀ █▀█ █▄▄ █░█ █▀█ █▄█ ██▄ ▄█
library(pacman)
p_load(tidyverse)
p_load(ggthemes)
p_load(glue)
p_load(onewaytests)
p_load(skimr)
p_load(knitr)
```

---

## Textbook Problems

```{r}
# helper function for generating tibble with important values 
report_values <- function(test, alpha) {
  test.method <- ifelse(
    str_detect(test$method, "Chi-squared"), "chi.squared", "anova")
  switch(test.method,
         "chi.squared" = tibble(
           alpha = alpha,
           degrees = unname(test$parameter),
           cv = round(qchisq(p = alpha, df = degrees, 
                             lower.tail = FALSE), 3),
           test.value = round(unname(test$statistic), 3)),
         "anova" = tibble(
           alpha = alpha,
           df.N = test$parameter[1],
           df.D = test$parameter[2],
           cv = round(qf(p = alpha, 
                         df1 = df.N, df2 = df.D, 
                         lower.tail = FALSE), 2),
           test.value = round(test$statistic, 2)))}

# helper function for making the decision
decision <- function(values) {
  significant <- values$test.value > values$cv
  ifelse(significant == TRUE, 
         glue("Reject the null hypothesis. 
              The test value of {values$test.value} 
              is greater than the critical value of {values$cv}"), 
         glue("Do not reject the null hypothesis. 
              The test value of {values$test.value}
              is less than the critical value of {values$cv}"))}

# helper function for summarizing the result
results <- function(values, hypotheses) {
  significant <- values$test.value > values$cv
  claim <- gsub(" (claim)", "", hypotheses$h1, fixed = TRUE)
  ifelse(significant == TRUE, 
         glue("There is sufficient evidence to support 
              the claim that {claim}."), 
         glue("There is not enough evidence to support 
              the claim that {claim}."))}
```

### Section 11-1

#### 6. Blood Types
>
A medical researcher wishes to see if hospital patients in a large hospital have the same blood type distribution as those in the general population. The distribution for the general population is as follows: type A, 20%; type B, 28%; type O, 36%; and type AB = 16%. He selects a random sample of 50 patients and finds the following: 12 have type A blood, 8 have type B, 24 have type O, and 6 have type AB blood.
>
At α = 0.10, can it be concluded that the distribution is the same as that of the general population?

###### Log the known information

```{r}
data <- tibble(
  E = c(0.20, 0.28, 0.36, 0.16),
  O = c(12, 8, 24, 6))
chi.sq <- chisq.test(x = data$O, p = data$E)
values <- report_values(test = chi.sq, alpha = 0.10)
```

###### State the hypotheses and identify the claim.

```{r}
hypotheses <- 
  tibble(
    h0 = "the distribution for the general population is: type A, 20%; type B, 28%; type O, 36%; and type AB = 16%",
    h1 = "the distribution differs from that of the general population (claim)")
```

`r kable(hypotheses)`

###### Find the critical value and test value.

`r kable(values)`

###### Make the decision.

`r decision(values)`

###### Summarize the results.

`r results(values, hypotheses)`

#### 8. On-Time Performance by Airline
>
According to the Bureau of Transportation Statistics, on-time performance by the airlines is described as follows:

| Action | % of Time |
|--------|-----------|
| On time | 70.8 |
| National Aviation System delay | 8.2 |
| Aircraft arriving late | 9.0 |
| Other (because of weather and other conditions) | 12.0 |

>
Records of 200 randomly selected flights for a major airline company showed that 125 planes were on time; 40 were delayed because of weather, 10 because of a National Aviation System delay, and the rest because of arriving late. At α = 0.05, do these results differ from the government’s statistics?

###### Log the known information

```{r}
data <- tibble(
  E = c(.708, .082, .09, .12),
  O = c(125, 10, 25, 40))

chi.sq <- chisq.test(x = data$O, p = data$E)
values <- report_values(test = chi.sq, alpha = 0.03)
```

###### State the hypotheses and identify the claim.

```{r}
hypotheses <- 
  tibble(
    h0 = "the performance of airlines is that 70.8% were on time, 12% were delayed for various reasons, 8.2% were delayed by the National Aviation System, and 9% were delayed by other aircraft arriving late",
    h1 = "the proportions of delays are different from those stated in the null hypothesis (claim)")
```

`r kable(hypotheses)`

###### Find the critical value and test value.

`r kable(values)`

###### Make the decision.

`r decision(values)`

###### Summarize the results.

`r results(values, hypotheses)`

### Section 11-2

#### 8. Ethnicity and Movie Admissions
>
Are movie admissions related to ethnicity? A 2014 study indicated the following numbers of admissions (in thousands) for two different years. At the 0.05 level of significance, can it be concluded that movie attendance by year was dependent upon ethnicity?

|    | Caucasian | Hispanic | African American | Other |
|----|-----------|----------|------------------|-------|
2013 | 724       | 335      | 174              | 107   |
2014 | 370       | 292      | 152              | 140   |

###### Log the known information

```{r}
data <- data.frame(
  Caucasion = c(724, 370),
  Hispanic = c(335, 292),
  African_American = c(174, 152),
  Other = c(107, 140))
rownames(data) <- c(2013, 2014)

chi.sq <- chisq.test(data)
values <- report_values(test = chi.sq, alpha = 0.05)
```

###### State the hypotheses and identify the claim.

```{r}
hypotheses <- 
  tibble(
    h0 = "movie attendance by year is independent of the ethnicity of the movie goers",
    h1 = "movie attendance by year is dependent upon the ethnicity of the movie goers (claim)")
```

`r kable(hypotheses)`

###### Find the critical value and test value.

`r kable(values)`

###### Make the decision.

`r decision(values)`

###### Summarize the results.

`r results(values, hypotheses)`

#### 10. Women in the Military
>
This table lists the numbers of officers and enlisted personnel for women in the military. At α = 0.05, is there sufficient evidence to conclude that a relationship exists between rank and branch of the Armed Forces?

| Action | Officers | Enlisted |
|--------|----------|----------|
| Army   | 10,791   | 62,491   |
| Navy   |7,816     |42,750    |
| Marine Corps|932  |9,525     |
| Air Force|11,819  |54,344    |

###### Log the known information

```{r}
data <- data.frame(
  Officers = c(10791, 7816, 932, 11819),
  Enlisted = c(62491, 42750, 9525, 54344)
  )
rownames(data) <- c("Army", "Navy", "Marine Corps", "Air Force")

chi.sq <- chisq.test(data)
values <- report_values(test = chi.sq, alpha = 0.05)
```

###### State the hypotheses and identify the claim.

```{r}
hypotheses <- 
  tibble(
    h0 = "the rank of women personnel is independent of the military branch of service",
    h1 = "the rank of women personnel is dependent upon the military branch of service (claim)")
```

`r kable(hypotheses)`

###### Find the critical value and test value.

`r kable(values)`

###### Make the decision.

`r decision(values)`

###### Summarize the results.

`r results(values, hypotheses)`

### Section 12-1

#### 8. Sodium Contents of Foods
>
The amount of sodium (in milligrams) in one serving for a random sample of three different kinds of foods is listed. At the 0.05 level of significance, is there sufficient evidence to conclude that a difference in mean sodium amounts exists among condiments, cereals, and desserts?

|Condiments|Cereals|Desserts|
|----------|-------|--------|
|270       |260    |100     |
|130       |220    |180     | 
|230       |290    |250     |
|180       |290    |250     |
|80        |200    |300     |
|70        |320    |360     |
|200       |140    |300     |
|          |       |160     |

###### Log the known information

```{r}
data <- bind_rows(
  tibble(Type = "Condiments", Sodium = c(270, 130, 230, 180, 80, 70, 200, NA)),
  tibble(Type = "Cereals", Sodium = c(260, 220, 290, 290, 200, 320, 140, NA)),
  tibble(Type = "Desserts", Sodium = c(100, 180, 250, 250, 300, 360, 300, 160)))

one.anova <- aov.test(Sodium ~ Type, data = data, alpha = 0.05)
values <- report_values(test = one.anova, alpha = one.anova$alpha)
```

###### State the hypotheses and identify the claim.

```{r}
hypotheses <- 
  tibble(
    h0 = "the mean amount of sodium in condiments, cereals, and desserts is the same",
    h1 = "at least one mean is different from the others (claim)")
```

`r kable(hypotheses)`

###### Find the critical value and test value.

`r kable(values)`

###### Make the decision.

`r decision(values)`

###### Summarize the results.

`r results(values, hypotheses)`

### Section 12-2

#### 10. Sales for Leading Companies
>
The sales in millions of dollars for a year of a sample of leading companies are shown. At α = 0.01, is there a significant difference in the means?

|Cereal|Chocolate Candy|Coffee|
|------|---------------|------|
|578   |311            |261   |
|320   |106            |185   |
|264   |109            |302   |
|249   |125            |689   |
|237   |173            |      |

###### Log the known information

```{r}
data <- bind_rows(
  tibble(Type = "Cereal", Sales.in.millions = c(578, 320, 264, 249, 237)),
  tibble(Type = "Chocolate Candy", Sales.in.millions = c(311, 106, 109, 125, 173)),
  tibble(Type = "Coffee", Sales.in.millions = c(261, 185, 302, 689, NA)))

one.anova <- aov.test(Sales.in.millions ~ Type, data = data, alpha = 0.01)
values <- report_values(test = one.anova, alpha = one.anova$alpha)
```

###### State the hypotheses and identify the claim.

```{r}
hypotheses <- 
  tibble(
    h0 = "the mean amount of sales for cereal, chocolate candy, and coffee is the same",
    h1 = "at least one mean is different from the others (claim)")
```

`r kable(hypotheses)`

###### Find the critical value and test value.

`r kable(values)`

###### Make the decision.

`r decision(values)`

###### Summarize the results.

`r results(values, hypotheses)`

#### 12. Per-Pupil Expenditures
>
The expenditures (in dollars) per pupil for states in three sections of the country are listed. Using α = 0.05, can you conclude that there is a difference in means?

|Eastern third|Middle third|Western third|
|-------------|------------|-------------|
|4946         |6149        |5282         |
|5953         |7451        |8605         |
|6202         |6000        |6528         |
|7243         |6479        |6911         |
|6113         |            |             |

###### Log the known information

```{r}
data <- bind_rows(
  tibble(Type = "Eastern third", Expenditures = c(4946, 5953, 6202, 7243, 6113)),
  tibble(Type = "Middle third", Expenditures = c(6149, 7451, 6000, 6479, NA)),
  tibble(Type = "Western third", Expenditures = c(5282, 8605, 6528, 6911, NA)))

one.anova <- aov.test(Expenditures ~ Type, data = data, alpha = 0.05)
values <- report_values(test = one.anova, alpha = one.anova$alpha)
```

###### State the hypotheses and identify the claim.

```{r}
hypotheses <- 
  tibble(
    h0 = "the mean expenditures (in dollars) per student for states in three sections of the country are the same",
    h1 = "at least one mean is different from the others (claim)")
```

`r kable(hypotheses)`

###### Find the critical value and test value.

`r kable(values)`

###### Make the decision.

`r decision(values)`

###### Summarize the results.

`r results(values, hypotheses)`

### Section 12-3

#### 10. Increasing Plant Growth
>
A gardening company is testing new ways to improve plant growth. Twelve plants are randomly selected and exposed to a combination of two factors, a “Grow-light” in two different strengths and a plant food supplement with different mineral supplements. After a number of days, the plants are measured for growth, and the results (in inches) are put into the appropriate boxes.

|             |Grow-light 1   |Grow-light 2 |
|-------------|---------------|-------------|
|Plant food A |9.2, 9.4, 8.9  |8.5, 9.2, 8.9|
|Plant food B |7.1, 7.2, 8.5  |5.5, 5.8, 7.6|

>
Can an interaction between the two factors be concluded? Is there a difference in mean growth with respect to light? With respect to plant food? Use α = 0.05.

###### Log the known information

```{r}
data <- bind_rows(
  tibble(Plant.food = "A", Grow.light = c(9.2, 9.4, 8.9), Type = c("GL1")),
  tibble(Plant.food = "A", Grow.light = c(8.5, 9.2, 8.9), Type = c("GL2")),
  tibble(Plant.food = "B", Grow.light = c(7.1, 7.2, 8.5), Type = c("GL1")),
  tibble(Plant.food = "B", Grow.light = c(5.5, 5.8, 7.6), Type = c("GL2")))
data$Plant.food <- factor(data$Plant.food)
data$Type <- factor(data$Type)

two.anova <- aov(Grow.light ~ Plant.food * Type, data = data)

values <- tibble(
  alpha = 0.05,
  df.N = summary(two.anova)[[1]][[1]][1],
  df.D = summary(two.anova)[[1]][[1]][4],
  Plant.food.ss = summary(two.anova)[[1]][[2]][1],
  Plant.food.F = summary(two.anova)[[1]][[4]][1],
  Grow.light.ss = summary(two.anova)[[1]][[2]][2],
  Grow.light.F = summary(two.anova)[[1]][[4]][2],
  Interaction.ss = summary(two.anova)[[1]][[2]][3],
  Interaction.F = summary(two.anova)[[1]][[4]][3],
  cv = round(qf(p = alpha, df1 = df.N, df2 = df.D, lower.tail = FALSE), 2))
```

###### State the hypotheses and identify the claim.

```{r}
hypotheses <- 
  tibble(
    h0.interaction = "There is no interaction between the strength of the grow light and the plant food supplement",
    h1.interaction = "There is an interaction between the strength of the grow light and the plant food supplement (claim)",
    h0.plant.food = "There is no difference between mean growth with respect to the type of plant food supplement",
    h1.plant.food = "There is a difference between mean growth with respect to the type of plant food supplement (claim)",
    h0.grow.light = "There is no difference between mean growth with respect to the strength of the grow light",
    h1.grow.light = "There is a difference between mean growth with respect to the strength of the grow light (claim)",
    )
```

`r kable(hypotheses)`

###### Find the critical value and test value.

`r kable(values)`

###### Make the decision and summarize the results

Since the F test value for plant food (24.56) is greater than the critical value (5.32), reject the null hypothesis for plant food. There is sufficient evidence to support the claim that there is a difference between mean growth with respect to the type of plant food supplement.

Since the F test value for grow light (3.68) is lower than the critical value (5.32) and since the F test value for an interaction between plant food and grow light (1.44) is lower than the critical value (5.32), do not reject the null hypotheses. There is not sufficient evidence to support that there is a difference between mean growth with respect to the strength of the grow light or that there is an interaction between the strength of the grow light and the plant food supplement.

---

## Baseball Dataset - "On Your Own"

```{r}
baseball <- read_csv("/Users/justin/Desktop/ALY 6015/Assignments/M2/chi-square-testing-and-anova/baseball.csv")

skim(baseball)

baseball %>% 
  ggplot +
  aes(Year, W) +
  geom_smooth(method = "lm", se = FALSE, size = 0.5, fullrange = TRUE) +
  geom_line(size = 0.8) +
  facet_wrap(~ Team) +
  labs(title = "Plot 1: Wins per year by team with blue line of best fit",
       y = "Count of wins") +
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))
```

Billy Beane and Paul DePodesta changed baseball history by using statistics to recruit undervalued players. They examined the relationships between stats such as on-base percentage (OBP) and slugging percentage (SLG) in relation to runs scored (RS) and playoffs (binary), coming up with statistical methods to predict great players who could be recruited on a low budget. This data set is an exercise in finding "truth" in data, and the goal of this assignment is to apply a Chi-Squared Goodness of Fit test to see if there is a difference in the number of wins per team by decade.

To give an overview of the dataset, there are 15 variables and 1232 rows. There are 13 numeric variables (discrete or continuous), and 2 character variables (Team and League - nominal). The completion rate for all variables except RankSeason, RankPlayoffs, OOBP, and OSLG is 100%. These four variables with missing data have between a 20 to 30% completion rate. Data was collected from years 1962 to 2012, and as this assignments concerns the number of wins, it should be noted that the mean number of wins (W) is 80.9 with a standard deviation of 11.46, a maximum value of 116, and a minimum value of 40. The number of games played per team by year (G) remains relatively constant throughout all years with a mean value of 161.92 and a standard deviation of 0.62, implying that this data set is well-balanced and the games played (G) should not bias wins (W).

In Plot 1, wins per team by year is displayed with lines of best fit. Some teams have little data (e.g., MIA/ANA), some teams have sporadic data (e.g., LAA), and generally teams with data spanning all years have near-flat slopes on the line of best fit. Newer teams such as WSN have a steep positive slope, but their newness discredits this finding. NYM and ATL visually appear to have the steepest positive slope among teams that have played since the 1960s. PIT visually appears to have the steepest negative slope among teams that have played since the 1960s.

### Chi-Square Goodness-of-Fit Test

```{r}
# Extract decade from year
baseball$Decade <- baseball$Year - (baseball$Year %% 10)
```

#### Baseball wins by decade
>
The baseball dataset contains teams' number of wins by decade. At α = 0.05 can it be concluded that there is a difference in the number of wins by decade?

###### Log the known information

```{r}
wins.per.decade <- baseball %>%
  select(Decade, W) %>%
  group_by(Decade) %>%
  summarize(O = sum(W)) %>%
  ungroup() %>%
  mutate(E = O / sum(O)) %>%
  t()

chi.sq <- chisq.test(x = wins.per.decade["O",], 
                     p = wins.per.decade["E",])
values <- report_values(test = chi.sq, alpha = 0.05)
```

###### State the hypotheses and identify the claim.

```{r}
hypotheses <- 
  tibble(
    h0 = "there is no difference in the number of wins by decade",
    h1 = "there is a difference in the number of wins by decade (claim)")
```

`r kable(hypotheses)`

###### Find the critical value and test value.

`r kable(values)`

###### Make the decision.

`r decision(values)`

###### Summarize the results.

`r results(values, hypotheses)`

###### Does comparing the critical value with the test value provide the same result as
comparing the p-value from R with the significance level?

```{r}
chi.sq
```

The decision not to reject the null hypothesis is the same when comparing the p-value (1) from R with the significance level (alpha = 0.05).

---

```{r}
crops <- read_csv("/Users/justin/Desktop/ALY 6015/Assignments/M2/chi-square-testing-and-anova/crop_data.csv")

# The block variable is superfluous -- it can  be dropped.
crops <- crops %>% select(-block)
```

```{r}
crops$fertilizer <- factor(crops$fertilizer)
crops$density <- factor(crops$density)

two.anova <- aov(yield ~ fertilizer * density, data = crops)

values <- tibble(
  alpha = 0.05,
  df.N = summary(two.anova)[[1]][[1]][1],
  df.D = summary(two.anova)[[1]][[1]][4],
  fertilizer.ss = summary(two.anova)[[1]][[2]][1],
  fertilizer.F = summary(two.anova)[[1]][[4]][1],
  density.ss = summary(two.anova)[[1]][[2]][2],
  density.F = summary(two.anova)[[1]][[4]][2],
  Interaction.ss = summary(two.anova)[[1]][[2]][3],
  Interaction.F = summary(two.anova)[[1]][[4]][3],
  cv = round(qf(p = alpha, df1 = df.N, df2 = df.D, lower.tail = FALSE), 2))
```

###### State the hypotheses and identify the claim.

```{r}
hypotheses <- 
  tibble(
    h0.interaction = "There is no interaction between the fertilizer used and plant density",
    h1.interaction = "There is an interaction between the fertilizer used and plant density (claim)",
    h0.fertilizer = "There is no difference between the fertilizer used with respect to the grain yield",
    h1.fertilizer = "There is a difference between the fertilizer used with respect to the grain yield (claim)",
    h0.density = "There is no difference between the plant density with respect to the grain yield",
    h1.density = "There is a difference between the plant density with respect to the grain yield (claim)",
    )
```

`r kable(hypotheses)`

###### Find the critical value and test value.

`r kable(values)`

###### Make the decision and summarize the results

Since the F test value for fertilizer used (9.001) is greater than the critical value (3.1), reject the null hypothesis for fertilizer. There is sufficient evidence to support the claim that there is a difference between the fertilizer used with respect to the grain yield.

Since the F test value for plant density (15.19) is greater than the critical value (3.1), reject the null hypothesis for density. There is sufficient evidence to support the claim that there is a difference between the plant density with respect to the grain yield.

Since the F test value for an interaction between fertilizer used and plant density (0.63) is lower than the critical value (3.1), do not reject the null hypothesis. There is not sufficient evidence to support the claim that there is an interaction between the fertilizer used and plant density.

## References
